```{r, echo = FALSE, eval = TRUE}
    sessionInfo()
```
---
title: "BIOSTAT-M280: Homework #4"
author: "Darrick Shen (UID:604946049) "
subtitle: Due Mar 16th @ 11:59PM
output: html_document
---

Our Apache Yarn cluster hosts the [flights](http://stat-computing.org/dataexpo/2009/the-data.html) data representing 123 million flights over 22 years. Read the [lecture notes](http://hua-zhou.github.io/teaching/biostatm280-2018winter/slides/12-sparklyr/sparklyr-flights.html) on how to access the Yarn cluster. Connect to the database using `sparklyr` and answer following questions. You can base your answers on a specific year or the whole data set.

1. Map the top 10 busiest airports. Size of dots should reflect the number of flights through that destination.  
Hint: You may find this tutorial on [Making Maps in R](http://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html) helpful.


    ```{r, echo = FALSE, eval = TRUE, include = FALSE}
    # Here we are installing and loading all packages required.
    
    if (!"sparklyr" %in% rownames(installed.packages()))  
      install.packages("sparklyr", repos = "http://cran.rstudio.com/")
    
    if (!"dplyr" %in% rownames(installed.packages()))  
      install.packages("dplyr", repos = "http://cran.rstudio.com/")
    
    if (!"ggplot2" %in% rownames(installed.packages()))  
      install.packages("ggplot2", repos = "http://cran.rstudio.com/")
    
    if (!"ggmap" %in% rownames(installed.packages()))  
      install.packages("ggmap", repos = "http://cran.rstudio.com/")

    if (!"maps" %in% rownames(installed.packages()))  
      install.packages("maps", repos = "http://cran.rstudio.com/")
    
    if (!"mapdata" %in% rownames(installed.packages()))  
      install.packages("mapdata", repos = "http://cran.rstudio.com/")
    
    if (!"ggrepel" %in% rownames(installed.packages()))  
      install.packages("ggrepel", repos = "http://cran.rstudio.com/")

    if (!"lubridate" %in% rownames(installed.packages()))  
      install.packages("lubridate", repos = "http://cran.rstudio.com/")

    library(sparklyr)
    library(dplyr)
    library(ggplot2)
    library(ggmap)
    library(maps)
    library(mapdata)
    library(ggrepel)
    library(lubridate)

    Sys.setenv(SPARK_HOME="/usr/lib/spark")
    config <- spark_config()
    sc <- spark_connect(master = "local", config = config)

    flights_tbl <- tbl(sc,'flights')
    airports_tbl <- tbl(sc, 'airports')
    airlines_tbl <- tbl(sc, 'airlines')

    #head(flights_tbl)
    #head(airports_tbl)
    #head(airlines_tbl)
```


```{r}
    #Map the top 10 busiest airports. Size of dots should reflect the number of 
    #flights through that destination.


    # Determine the n count of flight origins
    originCount <- flights_tbl %>%
    filter(year == 1991) %>%
    group_by(origin) %>%
    tally() %>%
    arrange(desc(n)) %>%
    collect() %>%
    head(15)
    # Determine the n count of flight destinations
    departureCount <- flights_tbl %>%
    filter(year == 1991) %>%
    group_by(dest) %>%
    tally() %>%
    arrange(desc(n)) %>%
    collect() %>%
    head(15)

   colnames(originCount) <- c("origin", "outgoing")
   colnames(departureCount) <- c("destination", "incoming")

  # Determine the "busiest" airports by summing incoming and outcoming flights.
  flightCount <-originCount %>%
  inner_join(departureCount, by = c("origin" = "destination")) %>%
  transmute(airport = origin, totalFlights = incoming + outgoing) %>%
  arrange(desc(totalFlights)) %>%
  head(10)

  # Location Retrieval.
topTenLocations <- flightCount %>%
  inner_join(airports_tbl, by = c("airport" = "faa"), copy = TRUE) %>%
  select(airport, totalFlights, lon, lat) %>%
  mutate(lat = as.double(lat),
         lon = as.double(lon),
         proportion = totalFlights/sum(totalFlights)) %>%
  collect()

  # Create compressed RDS  file.
#library(tidyverse)
##write_rds(topTenLocations, "hw4/busiestAirports.rds")

  # Create a plot of the US MAP.

usmap <- map_data("usa")
states <- map_data("state")

map <- get_map(location = "united states", zoom = 4)

busyUSports <- ggmap(map) +
  geom_point(data = topTenLocations,
             aes(x = lon, y = lat, size = proportion),
                 color = "black") +
  geom_label_repel(data = topTenLocations,
                   aes(x= lon, y = lat, label =  airport, size = proportion
                      ), min.segment.length = 0, size = 3.5) +
  ggtitle("Busiest US Airports (Highest # of flights in/out) in 1991")
busyUSports

```



2. Map the top 10 busiest direct routes. Size of lines should reflect the number of flights through that route.

```{r}
#Map the top 10 busiest direct routes.
#Size of lines should reflect the number of flights through that route.

topRoutes <-flights_tbl %>%
  filter(year == 2002) %>%
  group_by(origin, dest) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  collect() %>%
  head(10)

# Retrieve latitude and longitide of origin and dest

topTenOrigin<- topRoutes %>%
  inner_join(airports_tbl, by = c("origin" = "faa"), copy = TRUE) %>%
  select(origin, lat, lon, name,n) %>%
  mutate(latO = as.double(lat),
         lonO = as.double(lon)) %>%
  collect()

topTenDest <- topRoutes %>%
  inner_join(airports_tbl, by = c("dest" = "faa"), copy = TRUE) %>%
  select(origin, lat, lon, name,n) %>%
  mutate(latD = as.double(lat),
         lonD = as.double(lon)) %>%
  collect()

q2 <- bind_cols(topTenOrigin, topTenDest)

q2 <- q2 %>%
  mutate(scale = (n / max(n)) ^ 10)

 #  For the labels, we want to remove duplicate origins so that its not cluttered: 3 different ways to do so:
q2Sliced <- q2 %>%
  group_by(origin) %>%
  filter(row_number() == 1)

q2Sliced2 <- q2 %>%
  .[c(1:6, 8),]

q2Sliced3 <- q2 %>%
  slice(c(1:6,8))

### google map
map <- get_map(location = "united states", zoom = 4)
ggmap(map) +
  geom_point(data = q2,
             aes(x = lonO, y = latO),
             colour = "red", alpha = 0.2) +
  geom_point(data = q2,
             aes(x = lonD, y = latD),
             colour = "red", alpha = 0.2)+
  geom_curve(data = q2,
             aes(x = lonO, y = latO, xend = lonD, yend = latD),    
             arrow = arrow(angle = 15, ends = "first", length = unit(0.2, "cm"), type = "closed"),
             alpha = 0.7, curvature = 0.15, inherit.aes = TRUE) +
  scale_size_continuous(range=c(1,30)) +
  coord_cartesian()


#### simple map
unitedStates <- ggplot(data = states) +
geom_polygon(aes(x = long, y = lat, fill = region, group = group), color = "white") + 
coord_fixed(1.3) +
  geom_point(data = q2,
             aes(x = lonO, y = latO),
             colour = "red", alpha = 0.2) +
  geom_point(data = q2,
             aes(x = lonD, y = latD),
             colour = "red", alpha = 0.2)+
  geom_curve(data = q2,
             aes(x = lonO, y = latO, xend = lonD, yend = latD),    
             arrow = arrow(angle = 15, ends = "first", length = unit(0.5, "cm"), type = "closed"),
             alpha = 0.7, curvature = 0.15, size = q2$scale, inherit.aes = TRUE) +
  scale_size_continuous(range=c(1,30)) +
  coord_cartesian() +
  guides(fill=FALSE) +
  geom_label_repel(data = q2Sliced,
                   aes(x= lonO, y = latO, label = q2Sliced$origin, size = 2)
                  , min.segment.length = 0, size = 3.5) +
  ggtitle("Busiest US Routes in 2002")
unitedStates


```


0. LAX:

    (a). Reproduce above plot. Visualize and explain some prominent features you observe. For example, what happened at points 1-5?
    
```{r}
#(a). Reproduce above plot. Visualize and explain some prominent features you observe. For example, what happened at points 1-5?
laxFlights <-flights_tbl %>%
  filter(origin == "LAX" | dest == "LAX") %>%
  select(origin, dayofmonth, month, year) %>%
  collect()

#Subset Between 2008 and 1998 Years of LAX flights.
laxFlightDate <- laxFlights %>%
  mutate(date = make_date(year, month, dayofmonth)) %>%
  filter(date >= ymd(19980101) & date <= ymd(20081231)) %>%
  collect()

# Create compressed RDS file.
#write_rds(laxFlightDate, "hw4/laxflightdates.rds")
# Plot

laxFlightDate %>%
  group_by(date) %>%
  count() %>%
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  geom_label(aes(x = ymd(20011230), y = 1110, label = "1")) +
  geom_label(aes(x = ymd(20041201), y = 1000, label = "2")) +
  geom_label(aes(x = ymd(20040630), y = 950, label = "3")) +
  geom_label(aes(x = ymd(20080101), y = 1300, label = "4")) +
  geom_label(aes(x = ymd(20010130), y = 1150, label = "5")) +
  ggtitle("LAX air traffic") 

 # Visualize prominent features.

# Point 1
laxFlightDate  %>%
  group_by(date) %>%
  filter(date >= ymd(20010901) & date <= ymd(20011201)) %>%
  group_by(date) %>%
  count() %>%
  arrange(n) %>%
  collect() %>%
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  ggtitle("Point 1")
  
  

# Point 2
laxFlightDate  %>%
  group_by(date) %>%
  filter(date >= ymd(20040731) & date <= ymd(20050101)) %>%
  group_by(date) %>%
  count() %>%
  arrange(n) %>%
  collect() %>%
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  ggtitle("Point 2")


# Point 3
laxFlightDate  %>%
  group_by(date) %>%
  filter(date >= ymd(20040301) & date <= ymd(20041101)) %>%
  group_by(date) %>%
  count() %>%
  arrange(n) %>%
  collect() %>%
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  ggtitle("Point 3")

# Point 4
laxFlightDate  %>%
  group_by(date) %>%
  filter(date >= ymd(20070928) & date <= ymd(20080130)) %>%
  group_by(date) %>%
  count() %>%
  arrange(n) %>%
  collect() %>%
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  ggtitle("Point 4")

# Point 5
laxFlightDate  %>%
  group_by(date) %>%
  filter(date >= ymd(20070928) & date <= ymd(20080130)) %>%
  group_by(date) %>%
  count() %>%
  arrange(n) %>%
  collect() %>%
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  ggtitle("Point 5")
```

    (b). Visualize and explain seasonal effects.
```{r}

flightsBySeason <-flights_tbl %>%
  select(origin, dest, month, year) %>%
  filter(year >= 1998 && year <= 2008) %>%
  filter(dest == "LAX" | origin == "LAX") %>%
  mutate(season = if_else(month %in% c(3, 4, 5), "spring",
                          if_else(month %in% c(6, 7, 8), "summer",
                                  if_else(month %in% c(9,10, 11), "fall",
                                          if_else(month %in% c(12, 1, 2), "winter", "")
                                          )
                                  )
                          )
        ) %>%
  #group_by(year, season) %>%
  #tally() %>%
  count(year, season) %>%
  arrange(year) %>%
  collect()

ggplot(flightsBySeason) +
  geom_col(aes(x = as.factor(year), y = n, fill = season),
           color = "black") +
  labs(x = "Year", y = "Number of Flights", title = "LAX Air Traffic by Season") +
  scale_fill_brewer(palette = "Pastel2")
```
  
    (c). Visualize and explain weekly effects.
```{r}
flightsByWeekday <- flights_tbl %>%
  select(origin, dest, year, dayofweek) %>%
  filter(year >= 1998 && year <= 2008) %>%
  filter(dest == "LAX" | origin == "LAX") %>%
  count(year, dayofweek) %>%
  collect()

flightsByWeekday$dayofweek<-as.factor(flightsByWeekday$dayofweek)
ggplot(flightsByWeekday) +
  geom_col(aes(x = as.factor(year), y = n, fill = dayofweek),
           color = "black") +
  labs(x = "Year", y = "Number of Flights", title = "LAX Air Traffic by Day of Week") +
  scale_fill_brewer(palette = "Pastel2")


```
  
    (d). Map top 10 destinations from LAX. Size of dots should reflect the number of flights from LAX to that destination.
```{r}
# Filter the routes by destination: LAX.
topRoutes <-flights_tbl %>%
  select(origin, dest) %>%
  filter(origin == "LAX") %>%
  collect() %>% 
  count(dest) %>%
  arrange(desc(n)) %>%
  head(10)

# Location Retrieval.
destLAX <- topRoutes %>%
  left_join(airports_tbl, by = c("dest" = "faa"), copy = TRUE) %>%
  select(dest, lon, lat, n) %>%
  mutate(lat = as.double(lat),
         lon = as.double(lon)
         ) %>%
  mutate(airport = dest, proportion = n / sum(n))

ggmap(map) +
  geom_point(data = destLAX,
             aes(x = lon, y = lat, size = proportion),
             color = "black") +
  geom_label_repel(data = destLAX,
                   aes(x= lon, y = lat, label =  dest, size = proportion
                   ), min.segment.length = 0, size = 3.5) +
  labs(title = "Top 10 Locations from LAX")
```
0. Build a predictive model for the arrival delay (`arrdelay`) of flights flying from LAX. Use the same filtering criteria as in the [lecture notes](http://hua-zhou.github.io/teaching/biostatm280-2018winter/slides/12-sparklyr/sparklyr-flights.html) to construct training and validation sets. You are allowed to use a maximum of 5 predictors. The prediction performance of your model on the validation data set will be an important factor for grading this question.
```{r}
# Select data from an out of time sample
system.time(
modelFlights <- flights_tbl %>%
    filter(origin == "LAX") %>%
    filter(!is.na(arrdelay) & !is.na(depdelay) & !is.na(distance)) %>%
    filter(depdelay > 15 & depdelay < 240) %>%
    filter(arrdelay > -60 & arrdelay < 360) %>%
    filter(year >= 2003 & year <= 2007) %>%
    left_join(airlines_tbl, by = c("uniquecarrier" = "code")) %>%
    mutate(weekDay = as.character(dayofweek)) %>%
    select(year, month, arrdelay, depdelay, distance, uniquecarrier, description, weekDay)
)

sampleSplit <- modelFlights %>%
      sdf_partition(train = 0.8, valid = 0.2, seed= 250)

mlr <- sampleSplit$train %>%
  ml_linear_regression(arrdelay ~ distance + depdelay + weekDay)
summary(mlr)

# Apply MLR to LAX flight data in 2018.
system.time(
LAX2008 <-flights_tbl %>%
  filter(origin == "LAX") %>%
  filter(!is.na(arrdelay) & !is.na(depdelay) & !is.na(distance)) %>%
  filter(depdelay > 15 & depdelay < 240) %>%
  filter(arrdelay > -60 & arrdelay < 360) %>%
  filter(year == 2008) %>%
  left_join(airlines_tbl, by = c("uniquecarrier" = "code")) %>%
  mutate(weekDay = as.character(dayofweek)) %>%
  select(year, month, arrdelay, depdelay, distance, uniquecarrier, description, weekDay)
)

#Determine prediction model.

predictionVal2008 <- sdf_predict(LAX2008, mlr) %>%
  mutate(squareRes = (prediction - arrdelay)^2) %>%
  collect

# Calculate Root Mean Squared Error.
rootmse <- sqrt(sum(predictionVal2008$squareRes)/count(predictionVal2008))
rootmse

# Visualize actual arrival delay against predicted arrival delay.

PredictionData <- predictionVal2008 %>%
    group_by(weekDay) %>%
    summarize(actual = mean(arrdelay), prediction = mean(prediction), freq = n())

  # Plot
  ggplot(PredictionData, aes(actual, prediction)) + 
  geom_point(alpha = 0.75, color = 'red', shape = 3) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.15, color = 'blue') +
  geom_text(aes(label = substr(weekDay, 1, 20)), size = 3, alpha = 0.75, vjust = -1) +
  labs(title='Actual arrival Delay on Predicted Delay', x = 'Actual', y = 'Predicted')
```
    
0. Visualize and explain any other information you want to explore.


```{r}
  #Map the top 10 Origins to SFO in 1991.
  
  topSF <-flights_tbl %>%
    filter(year == 1991 & dest == "SFO") %>%
    group_by(origin, dest) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    collect() %>%
    head(10)
  
  # Retrieve latitude and longitide of origin and dest
  
  topTenSF <- topSF %>%
    inner_join(airports_tbl, by = c("origin" = "faa"), copy = TRUE) %>%
    select(origin, dest, lat, lon, name,n) %>%
    mutate(latO = as.double(lat),
           lonO = as.double(lon)) %>%
    collect()
  
  usmap <- map_data("usa")
  states <- map_data("state")
  SFOflights <- ggplot(data = states) +
    geom_polygon(aes(x = long, y = lat, fill = region, group = group), color = "white") + 
    coord_fixed(1.3) +
    geom_point(data = topTenSF,
               aes(x = lonO, y = latO, size = n),
               colour = "red", alpha = 0.3) +
    geom_point(data = topTenSF,
               aes(x = -122.375, y = 37.61899948120117),
               colour = "red", alpha = 0.2)+
    geom_curve(data = topTenSF,
               aes(x = lonO, y = latO, xend = -122.375, yend = 37.61899948120117),    
               arrow = arrow(angle = 15, length = unit(0.5, "cm"), type = "closed"),
               alpha = 0.7, inherit.aes = TRUE) +
    scale_size_continuous(range=c(1,30)) +
    coord_cartesian() +
    guides(fill = FALSE) +
    ggtitle("Top 10 Flight Origins to SFO in 1991")
  SFOflights
```

  