---
title: "hw2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = TRUE)
```

## 7.3.4 Exercises:
1. Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

    ```{r}
    library(ggplot2)
    ggplot(data = diamonds, mapping = aes(x = x)) +
      geom_freqpoly(binwidth = 0.01) +
      coord_cartesian(xlim = c(0,20)) -> p1

    ggplot(data = diamonds, mapping = aes(x = y)) +
      geom_freqpoly(binwidth = 0.01) +
      coord_cartesian(xlim = c(0,20)) -> p2

    ggplot(data = diamonds, mapping = aes(x = z)) +
      geom_freqpoly(binwidth = 0.01) +
      coord_cartesian(xlim = c(0,20)) -> p3

    source("http://peterhaschke.com/Code/multiplot.R")
    multiplot(p1, p2, p3, cols=1)
    ```
The plot for z shows measurements significantly shorter than those found in x and y, while x and y both look to be distributed very similarly. y and z both appear to have outliers. It makes sense to conclude that z is depth, while x is length, y is width (but x and y could be vice versa, as it is hard to distinguish between the two).

2. Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)


    ```{r}
    library(dplyr)

      # entire plot shows an unusual bump around ~8000.
    ggplot(data = diamonds, mapping = aes(x = price)) +
      geom_freqpoly(binwidth = 1) -> p1
    
      # zoom in on lower section and discover no price at ~1500.
    lower <- diamonds %>% 
      filter(price < 2000)
    ggplot(data = lower, mapping = aes(x = price)) +
      geom_histogram(binwidth = 10) -> p2
    
      # zoom in on mid-section.
    bigger <- diamonds %>%
      filter(price <10000 & price > 6000)
    ggplot(data = bigger, mapping = aes(x = price)) +
      geom_histogram(binwidth = 50) -> p3
    
      # zoom in on upper section.
    upper <- diamonds %>% 
      filter(price >5000)
    ggplot(data = upper, mapping = aes(x = price)) +
      geom_histogram(binwidth = 10) -> p4

    multiplot(p1, p2, p3, p4, cols=2)
```

We notice generally that there are exponentially more lower priced diamonds than expensive diamonds. The top left graph shows a slight increase in diamonds priced around ~8000. The lower left graph shows an unusual absence of diamonds priced at ~1500.


