title: "Biostat M280 Homework 3"
subtitle: Due Mar 2 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1 LA City Employee Payroll

The `/home/m280-data/la_payroll/LA_City_Employee_Payroll.csv` file on teaching server contains payroll information of LA City employees in years 2013-2017. It was downloaded from [LA City Controller's Office](https://controllerdata.lacity.org/Payroll/City-Employee-Payroll/pazn-qyym). Make a Shiny app to facilitate exploratory data analysis. 

1. For efficiency of the Shiny app, you should first pre-process, pare down, tidy, and save the data, e.g., as a compressed RDS file, to be used in the app.

0. **Total payroll by LA City**. Visualize the total LA City payroll of each year, with breakdown into base pay, overtime pay, and other pay.

0. **Who earned most?** Visualize the payroll information (total payment with breakdown into base pay, overtime pay, and other pay, Department, Job Title) of the top $n$ highest paid LA City employees in a specific year. User specifies $n$ (default 10) and year (default 2017).

0. **Which departments earn most?** Visualize the mean or median payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ earning departments. User specifies $n$ (default 5), year (default 2017), and method (mean or median, default median).

0. **Which departments cost most?** Visualize the total payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ expensive departments. User specifies $n$ (default 5) and year (default 2017).

0. Visualize any other information you are interested in.

0. Publish your Shiny app to <https://www.shinyapps.io> and share the link.

## Q2 LA City Parking War

The SQLite database `/home/m280-data/la_parking/LA_Parking_Citations.sqlite` on teaching server contains information about parking tickets in LA City. It was downloaded from [LA Open Data Portal](https://data.lacity.org/A-Well-Run-City/Parking-Citations/wjz9-h9np). Connect to the database and answer following questions using plots and summary statistics. In this exercise, you are **not** allowed to load whole data into memory. Use the _transform in database, plot in R_ strategy.

1. How many tickets are in this data set? Which time period do these tickets span? Which years have most data?
    ```{r}
        if (!"tidyverse" %in% rownames(installed.packages()))  
          install.packages("tidyverse", repos = "http://cran.rstudio.com/")
    
        if (!"DBI" %in% rownames(installed.packages()))  
          install.packages("DBI", repos = "http://cran.rstudio.com/")
    
        if (!"RSQLite" %in% rownames(installed.packages()))  
          install.packages("RSQLite", repos = "http://cran.rstudio.com/")
    
        if (!"dbplyr" %in% rownames(installed.packages()))  
          install.packages("dbplyr", repos = "http://cran.rstudio.com/")

        if (!"knitr" %in% rownames(installed.packages()))  
          install.packages("knitr", repos = "http://cran.rstudio.com/")
    
        library(knitr)
        library(tidyverse)
        library(DBI)
        library(dplyr)
        library(RSQLite)
    
          # connect to database.
        con <- dbConnect(RSQLite::SQLite(),
                         dbname = "/home/m280-data/la_parking/LA_Parking_Citations_Extra.sqlite")
        dbListTables(con)

          # querying tables.
        res <- dbSendQuery(con, "SELECT * FROM latix")
          #  dbFetch(res)
        latix_sql <- dplyr::tbl(con, "latix")
          # 4,044,488 rows
    ```
We can see from the total number of rows of the latix database, 4,044,448 rows represent the number of tickets, including those with a NA ID #
    ```{sql connection=con}
    SELECT
        count(DISTINCT Ticket_number)
    FROM
        latix
    ```
SQL code automatically removes the N/A ID results, leading to a 4,044,338 distinct ticket numbers.
    ```{r}
      # Here we are displaying the top ticketed times.
    ascend <- latix_sql %>%
      filter(!is.na(Issue_Year)) %>%
      select(Issue_Year, Issue_Month, Issue_Month,
             Issue_Day, Issue_Hour, Issue_Minute) %>%
      arrange(Issue_Year, Issue_Month, Issue_Month,
              Issue_Day, Issue_Hour, Issue_Minute)
    head(ascend)
      # Here we are displaying the lowest ticketed times.
    descend <- latix_sql %>%
    filter(!is.na(Issue_Year)) %>%
      select(Issue_Year, Issue_Month, Issue_Month,
             Issue_Day, Issue_Hour, Issue_Minute) %>%
      arrange(desc(Issue_Year), desc(Issue_Month), desc(Issue_Month),
              desc(Issue_Day), desc(Issue_Hour), desc(Issue_Minute))
    head(descend)

    latix_sql %>% 
      filter(!is.na(Issue_Year)) %>%
      group_by(Issue_Year) %>%
      summarize (n = n())

    latix_sql %>%
      group_by(Issue_Year) %>%
      filter(!is.na(Issue_Year)) %>%
      summarise(num=n()) %>%
      collect() %>%
      ggplot() +
        geom_col(aes(x = Issue_Year, y = num)) +
        scale_x_continuous("Issue Years") +
        ggtitle("Distribution of Issued Ticket")
```

There are 4044338 distinct ticket numbers. In addition 150 tickets had a NA ticket ID 
The oldest issued ticket was on 2010-04-27, at 21:40, while the most recently issued ticket was on 2017-12-30, at 01:41.
The most tickets were issued in 2015 (2,161,119 tickets). 

2) When (which hour, weekday, month day, and month) are you most likely to get a ticket and when are you least likely to get a ticket?

```{sql connection=con}
SELECT Issue_Hour, Issue_Day, Issue_Month, Issue_Wday, count(*) FROM latix
group by Issue_Hour, Issue_Day, Issue_Month, Issue_Wday
order by count (*) desc
limit 5;
```




    ```{r, echo = TRUE, eval = TRUE, message = FALSE}

    latix_sql %>%
      group_by(Issue_Hour) %>%
      filter(!is.na(Issue_Hour)) %>%
      summarise(num = n()) %>%
      collect() %>%
      ggplot() +
        geom_col(aes(x = Issue_Hour, y = num))
  
    latix_sql %>%
      group_by(Issue_Day) %>%
      filter(!is.na(Issue_Day)) %>%
      summarise(num = n()) %>%
      collect() %>%
      ggplot() +
        geom_col(aes(x = Issue_Day, y = num))
  
    latix_sql %>% 
      group_by(Issue_Wday) %>%
      filter(!is.na(Issue_Wday)) %>%
      summarise(num=n()) %>%
      collect() %>%
      ggplot() +
        geom_col(aes(x = Issue_Wday, y = num))  +
        scale_x_continuous("Issue Week Days",
                           breaks = c(1, 2, 3, 4, 5, 6, 7),
                                      label = c("Sun.", "Mon.",
                                                "Tues.", "Wed.",
                                                "Thurs.", "Fri.",
                                                "Sat.")) +
                             ggtitle("Distribution of Issued Tickets on Weekdays")
  
    latix_sql %>%
      group_by(Issue_Month) %>%
      filter(!is.na(Issue_Month)) %>%
      summarise(num=n()) %>%
      collect() %>%
      ggplot() +
        geom_col(aes(x = Issue_Month, y = num)) +
          scale_x_continuous("Issue Months",
                             breaks = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12), 
                             label = c("Jan.", "Feb.", "Mar.", "April", "May.",
                                       "June", "July", "Aug.", "Sept.", "Oct.",
                                       "Nov.", "Dec.")) +
        ggtitle("Distribution of Issued Ticket")
  
   latix_sql %>%
     filter(!is.na(Issue_Hour)) %>%
     count(Issue_Hour) %>%
     arrange(n) %>%
     collect() %>%
     kable()

    latix_sql %>%
      filter(!is.na(Issue_Day)) %>%
      count(Issue_Day) %>%
      arrange(n) %>%
      collect() %>%
      kable()

    latix_sql %>%
      filter(!is.na(Issue_Month)) %>%
      count(Issue_Month) %>%
      arrange(n) %>%
      collect() %>%
      kable()
    ```
The most likely hour is 12 (noon), whilethe least likely hour 5 (AM).
The most likely month day is 22, while the least likely month day is 31.
The most likely weekday is Tuesday, while the least likely weekday is  Saturday.
The most likely Month is March, while the least likely month is Novemeber.

3) Which car makes received most citations?
    ```{r}
    latix_sql %>%
      group_by(Make) %>%
        tally() %>%
        collect() %>%
        filter(n == max(n)) %>%
        arrange(desc(n))

    # plot of cars with most cars;
      latix_sql %>%
        group_by(Make) %>%
        summarise(num = n()) %>%
        arrange(desc(num)) %>%
        collect() %>%
        top_n(5)%>%
        ggplot() +
          geom_col(aes(x = Make, y = num))
  
      latix_sql %>%
        group_by(Make) %>%
        filter( Make == "TOYO") %>%
        summarise(num=n()) %>%
        arrange(desc(num)) %>%
        collect() %>%
          kable()
  
    latix_sql %>%
      group_by(Make) %>%
      filter(Make == "TOYO") %>%
      summarise(num=n()) %>%
      arrange(desc(num)) %>%
      collect() %>%
      ggplot() +
        geom_col(aes(x = Make, y = num))
```
Toyota (TOYT) recieved the most citations at 669,548. We also notice 39,180 "TOYO" makes, which is likely also Toyota vehicles.

4) How many different colors of cars were ticketed? Which color attracted most tickets?
    ```{sql connection=con}
    SELECT
        count(DISTINCT Color)
    FROM
        latix
    ```
There are 65 different colors of cars ticketed.
    ```{r}
    col <- latix_sql %>%
      group_by(Color) %>%
      tally() %>%
      collect() %>%
      filter(n == max(n)) %>%
      arrange(desc(n))
    col

    latix_sql %>%
      group_by(Color) %>%
      summarise(num = n()) %>%
      arrange(desc(num)) %>%
      collect() %>%
      top_n(5)%>%
      ggplot() +
        geom_col(aes(x = Color, y = num))
```
Black color vehicles attracted the most tickets, indicated by Color "BK"  having the most tickets (n=862,283).

5) What are the most common ticket types?
    ```{r}
    violationType <- latix_sql %>%
      group_by(Violation_Description) %>%
      tally() %>%
      collect() %>%
      filter(n == max(n)) %>%
      arrange(desc(n))
    violationType

  latix_sql %>%
    group_by(Violation_Description) %>%
    summarise(num = n()) %>%
    arrange(desc(num)) %>%
    collect() %>%
    top_n(5)%>%
    ggplot() +
      geom_col(aes(x = Violation_Description, y = num)) +
      xlab("Violation Description") + ylab("Number") +
      coord_flip()
```
The most common ticket type was "NO PARK/STREET CLEAN", with a total of 1,149,021 tickets.

6) How much money was collected on parking tickets in 2015 and 2016?
    ```{r}
    latix_sql %>%
      select(Issue_Year, Fine_amount) %>%
      filter(Issue_Year == 2016 | Issue_Year == 2015) %>%
      group_by(Issue_Year) %>%
      summarize(totalfine = sum(Fine_amount, na.rm = TRUE)) %>%
      kable(col.names = c("Issue Year ", "   Total Fine Amount ($)"))

    latix_sql %>%
      select(Issue_Year, Fine_amount) %>%
      filter(Issue_Year == 2016 | Issue_Year == 2015) %>%
      group_by(Issue_Year) %>%
      summarize(totalfine = sum(Fine_amount, na.rm = TRUE)) %>%
      collect() %>%
      ggplot() +
        geom_col(aes(x = Issue_Year, y = totalfine)) + 
        scale_x_continuous("Issue Years", breaks = c(1, 2),
                     label = c("2015", "2016")) +
                     ggtitle("Total Fines During 2015-2016") +
                     labs( y =  "Total Fine")
```
The total fine amount in 2015 was 151,006,794 dollars, and the total fine amount in 2016 was 123,236,136 dollars.

7) Visualize any other information you are interested in.
    ```{r}
    latix_sql %>%
      group_by(RP_State_Plate) %>%
      summarise(num = n()) %>%
      arrange(desc(num)) %>%
      collect() %>%
      top_n(7) %>%
      kable()
  
    latix_sql %>%
      group_by(RP_State_Plate) %>%
      summarise(num = n()) %>%
      arrange(desc(num)) %>%
      collect() %>%
      top_n(7) %>%
      ggplot() +
        geom_col(aes(x = RP_State_Plate, y = num)) +
        xlab("Ticketed License Plate State") +
        ylab("Number") 
```

California is the most frequently ticketed license plate at 3,766,908 times, with the most ticketed out-of-state licenses being Arizona, Texas, and Nevada, in descending order.
