title: "Biostat M280 Homework 3"
subtitle: Due Mar 2 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1 LA City Employee Payroll

The `/home/m280-data/la_payroll/LA_City_Employee_Payroll.csv` file on teaching server contains payroll information of LA City employees in years 2013-2017. It was downloaded from [LA City Controller's Office](https://controllerdata.lacity.org/Payroll/City-Employee-Payroll/pazn-qyym). Make a Shiny app to facilitate exploratory data analysis. 

1. For efficiency of the Shiny app, you should first pre-process, pare down, tidy, and save the data, e.g., as a compressed RDS file, to be used in the app.

0. **Total payroll by LA City**. Visualize the total LA City payroll of each year, with breakdown into base pay, overtime pay, and other pay.

0. **Who earned most?** Visualize the payroll information (total payment with breakdown into base pay, overtime pay, and other pay, Department, Job Title) of the top $n$ highest paid LA City employees in a specific year. User specifies $n$ (default 10) and year (default 2017).

0. **Which departments earn most?** Visualize the mean or median payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ earning departments. User specifies $n$ (default 5), year (default 2017), and method (mean or median, default median).

0. **Which departments cost most?** Visualize the total payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ expensive departments. User specifies $n$ (default 5) and year (default 2017).

0. Visualize any other information you are interested in.

0. Publish your Shiny app to <https://www.shinyapps.io> and share the link.

## Q2 LA City Parking War

The SQLite database `/home/m280-data/la_parking/LA_Parking_Citations.sqlite` on teaching server contains information about parking tickets in LA City. It was downloaded from [LA Open Data Portal](https://data.lacity.org/A-Well-Run-City/Parking-Citations/wjz9-h9np). Connect to the database and answer following questions using plots and summary statistics. In this exercise, you are **not** allowed to load whole data into memory. Use the _transform in database, plot in R_ strategy.

1. How many tickets are in this data set? Which time period do these tickets span? Which years have most data?
```{r}
    if (!"tidyverse" %in% rownames(installed.packages()))  
      install.packages("tidyverse", repos = "http://cran.rstudio.com/")
    
    if (!"DBI" %in% rownames(installed.packages()))  
      install.packages("DBI", repos = "http://cran.rstudio.com/")
    
    if (!"RSQLite" %in% rownames(installed.packages()))  
      install.packages("RSQLite", repos = "http://cran.rstudio.com/")
    
    if (!"dbplyr" %in% rownames(installed.packages()))  
      install.packages("dbplyr", repos = "http://cran.rstudio.com/")

    if (!"knitr" %in% rownames(installed.packages()))  
      install.packages("knitr", repos = "http://cran.rstudio.com/")
    

  # connect to database.
con <- dbConnect(RSQLite::SQLite(), dbname = "/home/m280-data/la_parking/LA_Parking_Citations_Extra.sqlite")
dbListTables(con)

  # querying tables.
res <- dbSendQuery(con, "SELECT * FROM latix")
dbFetch(res)

latix_sql <- dplyr::tbl(con, "latix")
library("tidyverse")
# 4,044,488 rows
```
Includes tickets with NA ID.
```{sql connection=con}
SELECT
    count(DISTINCT Ticket_number)
FROM
    latix
```
4044338 distinct ticket numbers.
```{r}
ascend <- latix_sql %>%
filter(!is.na(Issue_Year)) %>%
select(Issue_Year, Issue_Month, Issue_Month, Issue_Day, Issue_Hour, Issue_Minute) %>%
arrange(Issue_Year, Issue_Month, Issue_Month, Issue_Day, Issue_Hour, Issue_Minute)
head(ascend)

descend <- latix_sql %>%
filter(!is.na(Issue_Year)) %>%
select(Issue_Year, Issue_Month, Issue_Month, Issue_Day, Issue_Hour, Issue_Minute) %>%
arrange(desc(Issue_Year), desc(Issue_Month), desc(Issue_Month), desc(Issue_Day), desc(Issue_Hour), desc(Issue_Minute))
head(descend)

latix_sql %>% 
  filter(!is.na(Issue_Year)) %>%
  group_by(Issue_Year) %>%
  summarize (n = n())

latix_sql %>%
group_by(Issue_Year) %>%
filter(!is.na(Issue_Year)) %>%
summarise(num=n()) %>%
collect() %>%
  ggplot() +
  geom_col(aes(x = Issue_Year, y = num)) +
  scale_x_continuous("Issue Years") + ggtitle("Distribution of Issued Ticket")
```

There are 4044338 distinct ticket numbers. In addition 150 tickets had a NA ticket ID 
The oldest issued ticket was on 2010-04-27, at 21:40.
The most recently issued ticket was on 2017-12-30, at 01:41.
The most tickets were issued in 2015 (2161119 tickets). Removing for tickets with no issue year, but including tickets with ID: NA.

2) When (which hour, weekday, month day, and month) are you most likely to get a ticket and when are you least likely to get a ticket?

```{sql connection=con}
SELECT Issue_Hour, Issue_Day, Issue_Month, Issue_Wday, count(*) FROM latix
group by Issue_Hour, Issue_Day, Issue_Month, Issue_Wday
order by count (*) desc
limit 5;
```




```{r}

latix_sql %>%
group_by(Issue_Hour) %>%
filter(!is.na(Issue_Hour)) %>%
summarise(num=n()) %>%
  collect() %>%
  ggplot() +
      geom_col(aes(x = Issue_Hour, y = num))
  
       

latix_sql %>%
group_by(Issue_Day) %>%
filter(!is.na(Issue_Day)) %>%
summarise(num=n()) %>%
  collect() %>%
  ggplot() +
      geom_col(aes(x = Issue_Day, y = num))
  
latix_sql %>% 
group_by(Issue_Wday) %>%
filter(!is.na(Issue_Wday)) %>%
summarise(num=n()) %>%
  collect() %>%
  ggplot() +
      geom_col(aes(x = Issue_Wday, y = num))  +
  scale_x_continuous("Issue Week Days", breaks = c(1, 2, 3, 4, 5, 6, 7),
                     label = c("Sun.", "Mon.", "Tues.", "Wed.", "Thurs.", "Fri.", "Sat.")) + ggtitle("Distribution of Issued Tickets on Weekdays")
  
latix_sql %>%
group_by(Issue_Month) %>%
filter(!is.na(Issue_Month)) %>%
summarise(num=n()) %>%
collect() %>%
  ggplot() +
  geom_col(aes(x = Issue_Month, y = num)) +
  scale_x_continuous("Issue Months", breaks = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
                     label = c("Jan.", "Feb.", "Mar.", "April", "May.", "June", "July", "Aug.", "Sept.", "Oct.", "Nov.", "Dec.")) + ggtitle("Distribution of Issued Ticket")
  


latix_sql %>%
  filter(!is.na(Issue_Hour)) %>%
  count(Issue_Hour) %>%
  arrange(n) %>%
  collect() %>%
  kable()

latix_sql %>%
  filter(!is.na(Issue_Day)) %>%
  count(Issue_Day) %>%
  arrange(n) %>%
  collect() %>%
  kable()

latix_sql %>%
  filter(!is.na(Issue_Month)) %>%
  count(Issue_Month) %>%
  arrange(n) %>%
  collect() %>%
  kable()

```
Most likely hour 12, least likely hour 5.
Most likely day 22, least likely day 31.
Most likely weekday tuesdays.
Least likely weekday 
Most likely Month March.
Least likely Month Nov.
3) Which car makes received most citations?
```{r}

  latix_sql %>%
      group_by(Make) %>%
      tally() %>%
      collect() %>%
      filter(n == max(n)) %>%
      arrange(desc(n))


# plot of cares with most cars;
  latix_sql %>%
  group_by(Make) %>%
  summarise(num=n()) %>%
  arrange(desc(num)) %>%
  collect() %>%
    top_n(5)%>%
  ggplot() +
      geom_col(aes(x = Make, y = num))
  
  latix_sql %>%
  group_by(Make) %>%
  filter( Make == "TOYO") %>%
  summarise(num=n()) %>%
  arrange(desc(num)) %>%
  collect() %>%
    kable()
  
  latix_sql %>%
  group_by(Make) %>%
  filter( Make == "TOYO") %>%
  summarise(num=n()) %>%
  arrange(desc(num)) %>%
  collect() %>%
  ggplot() +
      geom_col(aes(x = Make, y = num))
```
TOYT recieved the most citations at 669548. Also 39180 TOYO.
4) How many different colors of cars were ticketed? Which color attracted most tickets?
```{sql connection=con}
SELECT
    count(DISTINCT Color)
FROM
    latix
```
65 different colors
```{r}
  col <- latix_sql %>%
      group_by(Color) %>%
      tally() %>%
      collect() %>%
      filter(n == max(n)) %>%
      arrange(desc(n))
col

  latix_sql %>%
  group_by(Color) %>%
  summarise(num=n()) %>%
  arrange(desc(num)) %>%
  collect() %>%
    top_n(5)%>%
  ggplot() +
      geom_col(aes(x = Color, y = num))
```
 Color "BK" most tickets. (n=862283)
5) What are the most common ticket types?

```{r}
  violationType <- latix_sql %>%
          group_by(Violation_Description) %>%
          tally() %>%
          collect() %>%
          filter(n == max(n)) %>%
          arrange(desc(n))
violationType


  latix_sql %>%
  group_by(Violation_Description) %>%
  summarise(num=n()) %>%
  arrange(desc(num)) %>%
  collect() %>%
    top_n(5)%>%
  ggplot() +
      geom_col(aes(x = Violation_Description, y = num)) +
      xlab("Violation Description") + ylab("Number") +
      coord_flip()

```
Most common ticket type was "NO PARK/STREET CLEAN", with a total of 1149021.
6) How much money was collected on parking tickets in 2015 and 2016?
```{r}
  latix_sql %>%
  select(Issue_Year, Fine_amount) %>%
  filter(Issue_Year == 2016 | Issue_Year == 2015) %>%
  group_by(Issue_Year) %>%
  summarize(totalfine = sum(Fine_amount, na.rm = TRUE)) %>% kable(col.names = c("Issue Year ", "   Total Fine Amount ($)"))


  latix_sql %>%
  select(Issue_Year, Fine_amount) %>%
  filter(Issue_Year == 2016 | Issue_Year == 2015) %>%
  group_by(Issue_Year) %>%
  summarize(totalfine = sum(Fine_amount, na.rm = TRUE)) %>%
  collect() %>%
      ggplot() +
      geom_col(aes(x = Issue_Year, y = totalfine)) + 
       scale_x_continuous("Issue Years", breaks = c(1, 2),
                     label = c("2015", "2016")) +
                     ggtitle("Total Fines During 2015-2016") +
                     labs( y =  "Total Fine")
    
```
Total fine amount in 2015 was 151006794.
Total fine amount in 2016 was 123236136.
7) Visualize any other information you are interested in.
```{r}

  latix_sql %>%
  group_by(RP_State_Plate) %>%
  summarise(num=n()) %>%
  arrange(desc(num)) %>%
  collect() %>%
    top_n(7)%>%kable()
  
  latix_sql %>%
  group_by(RP_State_Plate) %>%
  summarise(num=n()) %>%
  arrange(desc(num)) %>%
  collect() %>%
    top_n(7)%>%
  ggplot() +
      geom_col(aes(x = RP_State_Plate, y = num)) +
      xlab("Ticketed License Plate State") + ylab("Number") 

```

California is the most frequently ticketed license plate at 3766908 times, with the most ticketed out-of-state licenses as Arizona, Texas, and Nevada, in descending order.
